The evolution of gene expression in the term placenta of viviparous mammals

### 2015-09-24 manuscript/manuscript.Rnw
start working on loading results
fix transcriptome assembly and mapping section

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
%\usepackage{txfonts}
%\usepackage{fancyref}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=reading,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6.5in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\pagestyle{fancy}

@@

@@

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
library("data.table")
library("ggplot2")
library("reshape2")
@

@@

@@

@@

@@

\DLA{This isn't how these were analyzed anymore; fix this.}

The quality of reads was evaluated using FASTQC v. 0.9.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Following this, reads were trimmed based on quality in Galaxy using
the FASTQ Trimmer v. 1.0.0
\cite{Blankenberg.ea2010:ManipulationofFASTQdata}. All transcriptomes
were then aligned to reference genomes and assembled using the Tuxedo
suite of next-generation sequencing analysis tools. Annotated
reference genomes were downloaded from Ensembl v. 64
\cite{Flicek.ea2012:Ensembl2012}, including \textit{Homo sapiens}
GRCh37, \textit{Pan troglodytes} CHIMP2.1 (6X coverage), \textit{Canis
  familiaris} BROADD2 (7.6X), \textit{Bos taurus} UMD3.1 (9.5X),
\textit{Loxodonta africana} loxAfr3 (7X), and \textit{Monodelphis
  domestica} BROADO5 (6.8X). The genome of the closely-related Pan
troglodytes was used to map the transcriptome of Pan paniscus; the
separation between the two recognized Pan species is generally
perceived to be ~1 Mya (Prufer et al., 2012). Reads were aligned to
reference genomes using Bowtie v. 0.12.7\cite{Langmead2009} and TopHat
v. 1.3.3 \cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
Cufflinks v. 1.2.1 was then used to assemble transcripts and quantify
expression values using fragments per kilobase of exon per million
mapped reads (FPKM)
\cite{Trapnell.ea2010:Transcriptassemblyandquantification}. RNA-Seq
metrics were computed using Picard
(\url{http://picard.sourceforge.net/}). For comparison, we also
downloaded the placenta Illumina RNA-Seq transcriptome of a mouse from
\cite{Wang.ea2011:surveyfornovelimprinted}. Raw reads were mapped as
above using the Ensembl v. 64 annotation of the \textit{Mus musculus} (house
mouse) genome (NCBI m37).

@@

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using DAVID
\cite{Huang.ea2009:Systematicandintegrativeanalysis,Sherman.ea2007:DAVIDKnowledgebasegene-centereddatabase}.
DAVID ranks functional annotation clusters according to enrichment
score (ES); functional annotation clusters are considered
significantly enriched if ES>1.3, as this corresponds to p<0.05
\cite{Huang.ea2009:Systematicandintegrativeanalysis}. We executed
analyses with two gene lists per transcriptome: 1) annotated genes
with expression levels of the top 5\% of the frequency distribution of
FPKM and 2) the top 100 genes with the highest FPKM. Due to the
paucity of annotated gene ontology terms for some species, gene lists
were compared using gene ontology terms derived from the human genome.

@@

@@

The percentage of bases that mapped to Ensembl annotated mRNA ranged
from ~35\% in elephant to ~85\% in human, with percentages of bases
mapping to annotated protein-coding segments of genes ranging from
~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
ranged from ~43\% in elephant to 80\% in bonobo. The number of
annotated transcripts to which reads mapped in each respective genomes
ranged from 4119 (bonobo) to 6506 (human) for transcripts with FPKM>10
and 11689 (dog) to 30733 (human) for transcripts with FPKM>1. The
greater degree of mapping to annotated protein-coding sequences in
humans corresponds to the greater number of Ensembl annotated
sequences in the human genome compared to the other five reference
genomes (Flicek, 2012).

@@

The top 50 expressed genes in each genome analyzed are shown in
\DLA{Tables 2-7}, and all genes for each transcriptome are listed in
\DLA{Supplementary Tables X-XX} [These are large Excel files to be
included with the Supplemental Material]. In addition, specific genes
including placental hormones and hormone-related genes, galectins, and
insulin-like growth factor related genes are shown with FPKM values
side to side for all transcriptomes including the reanalysis of mouse
from \cite{Wang.ea2011:surveyfornovelimprinted} in \DLA{Table 8}.
Below we describe each transcriptome separately, specifically
concentrating on highly-expressed genes.

@@

The top 50 expressed genes as measured by FPKM in the term placenta of
the human are shown in Table 2 with a list of all genes in
Supplemental Table X. The gene with by far the highest FPKM value
(101962) in humans is chorionic somatomammotropin hormone 1 (CSH1),
followed by chorionic somatomammotropin hormone 2 (CSH2), and
chorionic somatomammotropin hormone-like 1 (CSHL1), all three of which
are members of a distinct expansion of the growth hormone gene family
within hominoids. Another member of the growth hormone gene family,
GH2 (FPKM=729.639), is also expressed to a high degree in the human
placenta (Table 2). Ten pregnancy-specific beta-1-glycoprotein genes
are highly-expressed (FPKM>100) in the term placenta of humans, with
PSG4 showing the greatest degree of expression (FPKM= 5201.31; Table
2). These genes are members of a hominoid-specific expansion of the
carcinoembryonic antigen (CEA) family and known to be expressed by the
syncytiotrophoblast (Lei et al., 1992; Rand and Cross, 2008). Genes
involved in the production of estrogen (CYP19A1) and progesterone
(HSD3B1) show exceptionally high expression in the human placenta
(FPKM = 4817.15 and 829.846 respectively), as well as CGA (FPKM=
4793.08), which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (Table 2). The genes producing the beta subunits
of the chorionic gonadotropins (CGB, CGB1, CGB2, CGB5, CGB7, CGB8) are
either not expressed (CGB8) or expressed modestly (FPKM≤20.826; Table
8). Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. HBA1, HBB, HBG2, HBA2), the maternally
imprinted long noncoding RNA gene H19, genes coding for proteins
required for growth and/or cell proliferation (ADAM12, KISS1, IGF2,
GPC3, CSF3R), anticoagulation (TFPI2), and glucose transport (SLC2A1)
among others (Table 2, Supplemental Table X)

@@

The top 50 expressed genes as measured by FPKM in the term placenta of
the bonobo are shown in Table 3 with a list of all genes in
Supplemental Table X. Expression in the bonobo placenta differs
somewhat from that of the human, as evidenced by the makeup of the top
50 genes (Table 3). The highest expression (FPKM=10149.1) in bonobo is
proteoglycan 2, bone marrow (PRG2), which is involved in insulin
growth factor signaling and angiogenesis during pregnancy (Weyer and
Glerup, 2011). PRG2 is also expressed at relatively high levels in the
human placenta (FPKM=250.675; Supplemental Table X), although at
comparatively much lower levels than the bonobo. A bonobo gene closely
related to the chorionic somatomammotropins (Q95MK7\_PANTR) is also
expressed at high levels (FPKM=1232.56; Table 3). Two other Pan genes
included within this family (ENSPTRG00000030805 which shows similarity
to GH2 and ENSPTRG00000034560) show expression levels of FPKM>89. Six
pregnancy-specific beta-1-glycoprotein genes were expressed in the
bonobo placenta with FPKM ranging from 20.885-167.961 (Supplemental
Table X). Placental expression of the genes involved in estrogen and
progesterone production (CYP19A1, HSD3B1 respectively) show lower
expression in the bonobo as compared to humans (CYP19A1 FPKM= 48.6877;
HSD3B1 FPKM= 54.5301; Table 8). In addition, CGA is expressed in the
bonobo placenta (FPKM=202.629), but at much lower levels than humans
as well (Table 8). Two Ensembl-annotated beta subunits of the
chorionic gonadotropin genes are expressed in bonobo, one related to
CGB and another related to CGB7; both are not highly expressed (Table
8). Among the other top 50 expressed genes, there is high expression
of ribosomal proteins (e.g., RPL18, RPL23), keratins (e.g., KRT7,
KRT19), heat-shock proteins (e.g., HSPA5, HSPA1A), serpins (SERPINE1,
SERPINE2), insulin growth factor 2 (IGF2) and the hormone
adrenomedullin (ADM).

@@

The top 50 expressed genes as measured by FPKM in the term placenta of
the dog are shown in Table 4 with a list of all genes in Supplemental
Table X. The gene with the highest expression (FPKM=48399.9) in the
dog placenta is the hormone relaxin (RLN1). Other highly expressed
genes among the top 50 (Table 4) include genes involved in the
cytoskeleton (TMSB10, TMSB4X), anticoagulation (TFPI2, PLA2G7), innate
immunity (LCN2, GM2A), lipid metabolism (APOE), inhibition of matrix
metalloproteinases (TIMP2, TIMP3), and growth (PGF, EGFR, HTRA1). The
gene LGALS1, which shows evidence of immunosupression during pregnancy
(Than et al., 2008), shows high expression in dog (FPKM= 1405.02). In
addition, the gene SSLP1, a gene expressed in the placentas of cow and
sheep but not present in primates (Ushizawa et al., 2009), is also
expressed at high levels (FPKM= 1853.98). There is no placental
expression of the estrogen synthase gene (CYP19A1) in dogs, however
expression of HSD3B1 is relatively high (FPKM=157.622).

@@

The top 50 expressed genes as measured by FPKM in the term placenta of
the dog are shown in Table 5 with a list of all genes in Supplemental
Table X. The gene with the highest expression (FPKM=12488.8) was the
miRNA gene bta-mir-2887-1. This gene occurs in several copies in the
cow genome, and it was unclear as to how many or to which of 12
bta-mir-2887 genes this mapped (Supplemental Table X). A previous
study indicated that this miRNA gene was expressed in the bovine ovary
(Hossain et al., 2009). Several members of the ruminant-specific
pregnancy-associated glycoprotein family show especially high
expression in the cow placenta (Table 4, Supplemental Table X)
including PAG2 (FPKM=7255.34), PAG10 (2076.42), BT.83780 (1793.72),
PAG11 (962.19), PAG12 (469.751), BT.88769 (408.584), PAG1 (301.61),
PAG4 (186.496), PAG6 (157.348), PAG8 (153.225), and PAG16 (103.573).
In addition, 14 other PAGs were expressed with FPKM<40 (Supplemental
Table X). Other genes with high expression include genes involved with
regulation of growth (IGF2, PHLDA2), imprinted genes (GNAS, DLK1,
PHLDA2), and CSH1, a gene that is a member of a ruminant specific
expansion of placental prolactins and not directly related to the gene
of the same name in hominids. Other ruminant-specific placental
prolactins also show high expression in the cow placenta including
PRP1 (632.766), PRP5 (396.832), PRP6 (345.296), PRP2 (195.771), in
addition to 7 other placental prolactins with expression of FPKM<51
(Table 5; Supplemental Table X). SSLP1 (also called SOLD1), a gene
expressed in the cotyledonary villi of the ruminiant placenta and
involved in cell proliferation (Ushizawa et al., 2009, 2010), also
shows evidence of high expression (FPKM=1001.45). There is some
placental expression of the estrogen synthase gene (CYP19A1) in cows
(75.1346), and comparatively greater expression of HSD3B1 (FPKM=
157.622).

@@

In order to identify major functional categories of genes expressed in
the individual transcriptomes, we undertook functional annotation
clustering analyses using DAVID (see Methods). Concentrating on genes
with relatively high expression values, we conducted DAVID analyses
using a list of genes with FPKM values included within the top 5\%
distribution of expressed genes. Functional annotation clusters from
all six placenta transcriptomes show enrichment in similar clusters of
genes (Supplemental Table X). Significant enrichment in genes (ES>1.3,
p>0.05) in the top 5\% of FPKM values in at least five transcriptomes
included clusters such as ribosome, vesicle, cytoskeleton,
glycoprotein, lysosome, annexins, hormone stimulus or response,
growth, calcium binding, cell migration, basement membrane,
antioxidant activity or oxidative stress, cell adhesion, and
mitochondrion among others (Supplemental Table X).

>>>>

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
\usepackage[plain]{fancyref}
\renewcommand*{\fref}[1]{\Fref{#1}}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=reading,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6.5in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\pagestyle{fancy}

@@

% Acronyms
\acrodef{EV}[EV]{Extracellular Vesicle}
\acrodef{SVM}{Support Vector Machine}
\acrodef{ROC}{Receiver Operating Characteristic}
\acrodef{KNN}{K Nearest Neighbors}
\acrodef{SRA}{Sequence Read Archive}
\acrodef{FPKM}{Fragments Per Kilobase of exon per Million reads \cite{Trapnell.ea2010:Transcriptassemblyandquantification}}

@@

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
@

@@

<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm")))>>=
for (a in paste0("../data/",c("combined_fpkm"))) {
    load(a)
}
@

@@

\DLA{Fix how the sequencing was done}

@@

For comparison, additional sequences of placenta in \textit{Mus
  musculus} \cite{Wang.ea2011:surveyfornovelimprinted} and
\textit{Homo sapiens}
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} were obtained
from the \ac{SRA}.

@@

The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (Reference of ---
in \fref{tab:ref_genome_aligned_species}) were assembled into a
reference using Trinity v2014-07-17
\cite{Haas.ea2013:Denovotranscriptsequence} and annotated using
Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against either
the \textit{Homo sapiens} or \textit{Mus musculus} transcriptome
(\fref{tab:ref_genome_aligned_species}).
% 
All transcriptomes were then aligned to the appropriate reference
genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM} were quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
%
The genome of the closely-related \textit{Pan troglodytes} was used as
a reference for \textit{Pan paniscus}. The separation between the two
recognized Pan species is generally perceived to be $\approx 1$ Mya
\cite{Prüfer.ea2012:bonobogenomecomparedwith}.

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Common Name        & Reference & Annotation \\
    \midrule
    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & Ensembl 80 \\
    \textit{Ateles fusciceps}     & Spider monkey      & ---        & \textit{Homo sapiens} Ensembl 80 \\
    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & Ensembl 80 \\
    \textit{Bos taurus}           & Cow                & UMD3.1     & Ensembl 80 \\
    \textit{Canis familiaris}     & Dog                & CanFam3.1  & Ensembl 80 \\
    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & Ensembl 80 \\
    \textit{Homo sapiens}         & Human              & GRCh38     & Ensembl 80 \\
    \textit{Mus musculus}         & Mouse              & GRCm38     & Ensembl 80 \\
    \textit{Nannospalax galili}   & Blind mole rat     & ----       & \textit{Mus musculus} Ensembl 80 \\
    \textit{Monodelphis domestica} & Opossum           & BROAD05    & Ensembl 80 \\
    \textit{Spalax carmeli}       & Blind mole rat     & ----       & \textit{Mus musculus} Ensembl 80 \\
    \bottomrule
  \end{tabular}
  \caption{Reference genomes for aligned species}
  \label{tab:ref_genome_aligned_species}
\end{table}

@@

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using DAVID
\cite{Huang.ea2009:Systematicandintegrativeanalysis,Sherman.ea2007:DAVIDKnowledgebasegene-centereddatabase}.
DAVID ranks functional annotation clusters according to enrichment
score (ES); functional annotation clusters are considered
significantly enriched if ES>1.3, as this corresponds to p<0.05
\cite{Huang.ea2009:Systematicandintegrativeanalysis}. We executed
analyses with two gene lists per transcriptome: 1) annotated genes
with expression levels of the top 5\% of the frequency distribution of
\ac{FPKM} and 2) the top 100 genes with the highest \ac{FPKM}. Due to the
paucity of annotated gene ontology terms for some species, gene lists
were compared using gene ontology terms derived from the human genome.

@@

<<transcriptome.statistics>>=
@

@@

The percentage of bases that mapped to Ensembl annotated mRNA ranged
from ~35\% in elephant to ~85\% in human, with percentages of bases
mapping to annotated protein-coding segments of genes ranging from
~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
ranged from ~43\% in elephant to 80\% in bonobo. The number of
annotated transcripts to which reads mapped in each respective genomes
ranged from 4119 (bonobo) to 6506 (human) for transcripts with \ac{FPKM}>10
and 11689 (dog) to 30733 (human) for transcripts with \ac{FPKM}>1. The
greater degree of mapping to annotated protein-coding sequences in
humans corresponds to the greater number of Ensembl annotated
sequences in the human genome compared to the other five reference
genomes (Flicek, 2012).

@@

The top 50 expressed genes in each genome analyzed are shown in
\DLA{Tables 2-7}, and all genes for each transcriptome are listed in
\DLA{Supplementary Tables X-XX} [These are large Excel files to be
included with the Supplemental Material]. In addition, specific genes
including placental hormones and hormone-related genes, galectins, and
insulin-like growth factor related genes are shown with \ac{FPKM} values
side to side for all transcriptomes including the reanalysis of mouse
from \cite{Wang.ea2011:surveyfornovelimprinted} in \DLA{Table 8}.
Below we describe each transcriptome separately, specifically
concentrating on highly-expressed genes.

<<top.genes.function,echo=FALSE,results="hide">>=
top.genes.table <- function(species.name,n=50) {
    temp <- combined.fpkm[species==species.name &
                      !grepl("^MT-?",gene_short_name)][order(-mean_fpkm)][!duplicated(gene_short_name)][1:n,list(gene_short_name,human_name,mean_fpkm)]
    setnames(temp,c("Gene name","Human Name","FPKM"))
    if (species.name=="homo sapiens") {
        temp <- temp[,-2,with=FALSE]
    }
    return(temp)
}
@

@@

\begin{wraptable}{I}{0.3\textwidth}
  \tiny
<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
top.human.genes <- top.genes.table(species.name="homo sapiens")
print(xtable(top.human.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.human.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Homo sapiens}}
\label{tab:top50_human_placenta}
\end{wraptable}

The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
the human are shown in \fref{tab:top50_human_placenta} with a list of all genes in
Supplemental Table X. The gene with by far the highest \ac{FPKM} value
(101962) in humans is chorionic somatomammotropin hormone 1 (CSH1),
followed by chorionic somatomammotropin hormone 2 (CSH2), and
chorionic somatomammotropin hormone-like 1 (CSHL1), all three of which
are members of a distinct expansion of the growth hormone gene family
within hominoids. Another member of the growth hormone gene family,
GH2 (\ac{FPKM}=729.639), is also expressed to a high degree in the human
placenta (Table 2). Ten pregnancy-specific beta-1-glycoprotein genes
are highly-expressed (FPKM>100) in the term placenta of humans, with
PSG4 showing the greatest degree of expression (\ac{FPKM}= 5201.31; Table
2). These genes are members of a hominoid-specific expansion of the
carcinoembryonic antigen (CEA) family and known to be expressed by the
syncytiotrophoblast (Lei et al., 1992; Rand and Cross, 2008). Genes
involved in the production of estrogen (CYP19A1) and progesterone
(HSD3B1) show exceptionally high expression in the human placenta
(\ac{FPKM} = 4817.15 and 829.846 respectively), as well as CGA (\ac{FPKM}=
4793.08), which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (Table 2). The genes producing the beta subunits
of the chorionic gonadotropins (CGB, CGB1, CGB2, CGB5, CGB7, CGB8) are
either not expressed (CGB8) or expressed modestly (FPKM≤20.826; Table
8). Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. HBA1, HBB, HBG2, HBA2), the maternally
imprinted long noncoding RNA gene H19, genes coding for proteins
required for growth and/or cell proliferation (ADAM12, KISS1, IGF2,
GPC3, CSF3R), anticoagulation (TFPI2), and glucose transport (SLC2A1)
among others (Table 2, Supplemental Table X)

@@

\begin{wraptable}{O}{0.3\textwidth}
  \tiny
<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
top.human.genes <- top.genes.table(species.name="pan paniscus")
print(xtable(top.human.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.human.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Pan paniscus}}
\label{tab:top50_bonobo_placenta}
\end{wraptable}

The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
the bonobo are shown in Table 3 with a list of all genes in
Supplemental Table X. Expression in the bonobo placenta differs
somewhat from that of the human, as evidenced by the makeup of the top
50 genes (Table 3). The highest expression (\ac{FPKM}=10149.1) in bonobo is
proteoglycan 2, bone marrow (PRG2), which is involved in insulin
growth factor signaling and angiogenesis during pregnancy (Weyer and
Glerup, 2011). PRG2 is also expressed at relatively high levels in the
human placenta (\ac{FPKM}=250.675; Supplemental Table X), although at
comparatively much lower levels than the bonobo. A bonobo gene closely
related to the chorionic somatomammotropins (Q95MK7\_PANTR) is also
expressed at high levels (\ac{FPKM}=1232.56; Table 3). Two other Pan genes
included within this family (ENSPTRG00000030805 which shows similarity
to GH2 and ENSPTRG00000034560) show expression levels of \ac{FPKM}>89. Six
pregnancy-specific beta-1-glycoprotein genes were expressed in the
bonobo placenta with \ac{FPKM} ranging from 20.885-167.961 (Supplemental
Table X). Placental expression of the genes involved in estrogen and
progesterone production (CYP19A1, HSD3B1 respectively) show lower
expression in the bonobo as compared to humans (CYP19A1 \ac{FPKM}= 48.6877;
HSD3B1 \ac{FPKM}= 54.5301; Table 8). In addition, CGA is expressed in the
bonobo placenta (\ac{FPKM}=202.629), but at much lower levels than humans
as well (Table 8). Two Ensembl-annotated beta subunits of the
chorionic gonadotropin genes are expressed in bonobo, one related to
CGB and another related to CGB7; both are not highly expressed (Table
8). Among the other top 50 expressed genes, there is high expression
of ribosomal proteins (e.g., RPL18, RPL23), keratins (e.g., KRT7,
KRT19), heat-shock proteins (e.g., HSPA5, HSPA1A), serpins (SERPINE1,
SERPINE2), insulin growth factor 2 (IGF2) and the hormone
adrenomedullin (ADM).

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
the dog are shown in Table 4 with a list of all genes in Supplemental
Table X. The gene with the highest expression (\ac{FPKM}=48399.9) in the
dog placenta is the hormone relaxin (RLN1). Other highly expressed
genes among the top 50 (Table 4) include genes involved in the
cytoskeleton (TMSB10, TMSB4X), anticoagulation (TFPI2, PLA2G7), innate
immunity (LCN2, GM2A), lipid metabolism (APOE), inhibition of matrix
metalloproteinases (TIMP2, TIMP3), and growth (PGF, EGFR, HTRA1). The
gene LGALS1, which shows evidence of immunosupression during pregnancy
(Than et al., 2008), shows high expression in dog (\ac{FPKM}= 1405.02). In
addition, the gene SSLP1, a gene expressed in the placentas of cow and
sheep but not present in primates (Ushizawa et al., 2009), is also
expressed at high levels (\ac{FPKM}= 1853.98). There is no placental
expression of the estrogen synthase gene (CYP19A1) in dogs, however
expression of HSD3B1 is relatively high (\ac{FPKM}=157.622).

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
the dog are shown in Table 5 with a list of all genes in Supplemental
Table X. The gene with the highest expression (\ac{FPKM}=12488.8) was the
miRNA gene bta-mir-2887-1. This gene occurs in several copies in the
cow genome, and it was unclear as to how many or to which of 12
bta-mir-2887 genes this mapped (Supplemental Table X). A previous
study indicated that this miRNA gene was expressed in the bovine ovary
(Hossain et al., 2009). Several members of the ruminant-specific
pregnancy-associated glycoprotein family show especially high
expression in the cow placenta (Table 4, Supplemental Table X)
including PAG2 (\ac{FPKM}=7255.34), PAG10 (2076.42), BT.83780 (1793.72),
PAG11 (962.19), PAG12 (469.751), BT.88769 (408.584), PAG1 (301.61),
PAG4 (186.496), PAG6 (157.348), PAG8 (153.225), and PAG16 (103.573).
In addition, 14 other PAGs were expressed with \ac{FPKM}<40 (Supplemental
Table X). Other genes with high expression include genes involved with
regulation of growth (IGF2, PHLDA2), imprinted genes (GNAS, DLK1,
PHLDA2), and CSH1, a gene that is a member of a ruminant specific
expansion of placental prolactins and not directly related to the gene
of the same name in hominids. Other ruminant-specific placental
prolactins also show high expression in the cow placenta including
PRP1 (632.766), PRP5 (396.832), PRP6 (345.296), PRP2 (195.771), in
addition to 7 other placental prolactins with expression of \ac{FPKM}<51
(Table 5; Supplemental Table X). SSLP1 (also called SOLD1), a gene
expressed in the cotyledonary villi of the ruminiant placenta and
involved in cell proliferation (Ushizawa et al., 2009, 2010), also
shows evidence of high expression (\ac{FPKM}=1001.45). There is some
placental expression of the estrogen synthase gene (CYP19A1) in cows
(75.1346), and comparatively greater expression of HSD3B1 (\ac{FPKM}=
157.622).

@@

In order to identify major functional categories of genes expressed in
the individual transcriptomes, we undertook functional annotation
clustering analyses using DAVID (see Methods). Concentrating on genes
with relatively high expression values, we conducted DAVID analyses
using a list of genes with \ac{FPKM} values included within the top 5\%
distribution of expressed genes. Functional annotation clusters from
all six placenta transcriptomes show enrichment in similar clusters of
genes (Supplemental Table X). Significant enrichment in genes (ES>1.3,
p>0.05) in the top 5\% of \ac{FPKM} values in at least five transcriptomes
included clusters such as ribosome, vesicle, cytoskeleton,
glycoprotein, lysosome, annexins, hormone stimulus or response,
growth, calcium binding, cell migration, basement membrane,
antioxidant activity or oxidative stress, cell adhesion, and
mitochondrion among others (Supplemental Table X).
